Missing the Point: Non-Convergence in Iterative Imputation Algorithms

### 2020-05-11 Writeup/2.Manuscript.Rmd
Update research archive conform requirements

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width = 4.9, echo = FALSE, fig.align = "center") 
# knitr::opts_chunk$set(fig.width = 4.8, fig.height = 3.6, echo = FALSE, fig.align = "center")

# packages
library(mice)
library(dplyr)
library(ggplot2)
library(patchwork)

# load example non-convergence and figures
load("../Results/example_nonconv.Rdata")
load("../Results/example_conv.Rdata")
load("../Results/example_diagnostics.Rdata")
source("../3.PlotExampleNon-convergence.R")

# load results and figures
load("../Results/complete.Rdata") 
results <- results %>% filter(t<51)
source("../4.PlotResults.R")

# default graphing behavior
theme_update(
 plot.title = element_text(hjust = 0.5),
 plot.subtitle = element_text(hjust = 0.5),
 panel.border = element_blank(),
 panel.grid.major = element_blank(),
 panel.grid.minor = element_blank(),
 panel.background = element_blank(),
 axis.line = element_line(colour = "black"),
 legend.key = element_blank(),
 legend.position = "bottom",
 legend.margin = margin(0,0,0,0),
 # adjust text size
 text = element_text(size = 8),
 title = element_text(size = 8),
 legend.title = element_text(size = 7)
)
```

<!-- # Overview of sections -->

<!-- Intro -->

<!-- - This is missingness -->

<!-- - This is how we solve missingness -->

<!-- - For that we need an algorithm, but what if it has not converged? -->

<!-- - That is what we solve in this paper -->

<!-- Algorithmic convergence -->

<!-- - What is algorithmic convergence? -->

<!-- - When is it converged? Stationarity versus mixing. -->

<!-- - How is convergence currently investigated? And how should it be investigated? Multivariate versus univariate. -->

<!-- Methods -->

<!-- - Which methods are there? And how do they relate to mixing and stationarity? -->

<!-- - $\widehat{R}$ -->

<!-- - AC -->

<!-- - We can use these on any $\theta$ -->

<!-- Then Simulation study, Results, Discussion. -->

@@

Figure \ref{fig:diagram} provides an overview of the steps involved with MI. The missing part $y_{\rm mis}$ of an incomplete dataset is imputed $m$ times. This creates $m$ sets of imputed data $\dot y_{\rm imp, \ell}$, where $\ell = 1, 2, ..., m$. The imputed data is then combined with the observed data $y_{\rm obs}$ to create $m$ completed datasets. On each of these datasets the analysis of scientific interest is performed to estimate $Q$: the quantity of scientific interest (e.g., a regression coefficient). Since $Q$ is estimated on each completed dataset, $m$ separate $\hat{Q}_\ell$-values are obtained. Finally, the $\hat{Q}_\ell$-values are combined into a single pooled estimate $\bar{Q}$. The premise of multiple imputation is that $\bar{Q}$ is an unbiased and confidence-valid estimate of the true---but unobserved---scientific estimand $Q$ [@rubin96].

<!-- $Q$ is quantity of scientific interest: "A scientific estimand $Q$ is a quantity of scientific interest that we can calculate if we would observe the entire population" [@buur18, par 2.3.1] -->

```{r diagram, out.width="\\linewidth", include=TRUE, fig.cap="Scheme of the main steps in multiple imputation---from an incomplete dataset, to $m=3$ multiply imputed datasets, to $m=3$ estimated quantities of scientific interest $\\hat{Q}$, to a single pooled estimate $\\bar{Q}$."}
knitr::include_graphics("../Figures/DiagramMI.pdf") #adapted from [@buur18] par. 1.4.1.
```

A popular method to obtain imputations is to use the 'Multiple Imputation by Chained Equations' algorithm, shorthand 'MICE'[@mice]. With MICE, imputed values are drawn from the posterior predictive distribution of the missing values. The algorithm is named after its iterative nature: a multivariate distribution is obtained by iterating over a sequence of univariate imputations. Iteration, however, also introduces a potential threat to the validity of the imputations: What if the algorithm has not converged? Are the imputations then to be trusted? And can we rely on the inference obtained on the completed data? 
<!-- $\dot y_{\rm imp, \ell}$ $y_{\rm mis}$ -->

@@

<!-- The results of this study are guidelines for identifying non-convergence in iterative imputation algorithms, which will aid applied researchers in drawing valid inference from incomplete datasets. -->

<!-- **[Include sub-questions and make structure of the paper clear!]** How can non-convergence be identified diagnostically? Are common MCMC non-convergence diagnostics appropriate for MICE? And if so, which threshold should be used to diagnose non-convergence? How many iterations are sufficient/needed to be able to diagnose non-convergence? Are the default number of iterations sufficient (i.e., 5 in mice, 10 in SPSS and Stata, 30 in mi)? How severe is it when the algorithm has not converged? And what are guidelines for practice? Can the parameter of interest $Q$ be correct when the algorithm is not (yet) converged, and vice versa? -->

<!-- The defaults of common MI software: SPSS = 10 [(link)](https://www.ibm.com/support/knowledgecenter/SSLVMB_24.0.0/spss/mva/syn_multiple_imputation_impute.html), Mplus = 100?? [(link)](https://pdfs.semanticscholar.org/e20e/29e008592cbfbaa567931f74cdfdb5451405.pdf?_ga=2.55354671.54033656.1584698748-527613517.1584698748), Stata = 10 [(link)](https://www.stata.com/manuals13/mi.pdf, p. 139), Amelia = NA, because re-sampling, not convergence [(link)](https://cran.r-project.org/web/packages/Amelia/Amelia.pdf), en MI = 30 [(link)](https://cran.r-project.org/web/packages/mi/mi.pdf). -->

<!-- The goal of this research project is to develop novel methodology and guidelines for evaluating MI methods. These tools will subsequently be implemented in an interactive evaluation framework for multiple imputation, which will aid applied researchers in drawing valid inference from incomplete datasets. -->

@@

Let $y$ denote an $n \times k$ matrix containing the data values on $k$ variables for all $n$ units in a sample. The data value of unit $i$ ($i = 1, 2, \dots, n$) on variable $j$ ($j = 1, 2, \dots, k$) may be either observed or missing. The number of units $i$ with at least one missing value, divided by the total number of units $n$, is called the proportion of incomplete cases $p_{\rm inc}$ in dataset $y$. The collection of observed data values in $y$ is denoted by $y_{\rm obs}$; the missing part of $y$ is referred to as $y_{\rm mis}$. For each datapoint in $y_{\rm mis}$, we sample $m \times T$ plausible values, where $m$ is the number of imputations ($\ell = 1, 2, \dots, m$) and $T$ is the number of iterations in the imputation algorithm ($t = 1, 2, \dots, T$). The state-space of the algorithm at a certain iteration $t$ may be summarized by a scalar summary $\theta$ (e.g., the average of the imputed values). The collection of $\theta$-values between $t=1$ (summarizing the state-space of the algorithm at initialization) and $t=T$ (summarizing the state-space for the imputed values) will be referred to as an 'imputation chain'. 
<!-- **[If not defined elsewhere, add: $\theta$s are scalar summaries of the algorithm at a certain iteration (e.g., chain means; the average of the imputed values in each imputation chain).] ** -->

@@

<!-- When we declare approximate convergence, we are actually concluding that each individual sequence appears stationary and that the observed sequences -->
<!-- have mixed well with each other. @gelm13, p. 288-->

<!-- **The goal of imputation through e.g. MICE is to obtain a multivariate distribution by iterating over a sequence of univariate imputations. Traditionally, the inspection of the multivariate convergence is done by visually inspecting the univariate chain means and variances over the iterations. The downside to these approaches is that they either focus on the univariate state-space, or primarily track the change over the iterations of a multivariate outcome conform the scientific model of interest. Focusing on the convergence of these outcome parameters may influence this procedure in the sense that the model of evaluation favors the model of interest. Ideally, one would like to evaluate a model-independent parameter that summarizes the multivariate nature of the data. We propose eigenvalue as such a parameter, because it.... ** -->

@@

<!-- , since systematically higher imputed values lead to bias, and under-estimation of the variance within imputation chains poses a threat to confidence-validity. -->

@@

Currently, the recommended practice for evaluating the convergence of the MICE algorithm is through visual inspection. After running the imputation algorithm for a certain number of iterations, researchers are encouraged to produce traceplots. In a traceplot, a scalar summary of the state-space of the algorithm $\theta$ is plotted against the iteration number, as depicted in Figure \ref{fig:non-conv}. The default scalar summaries to inspect for the MICE algorithm are chain means and chain variances. Non-convergence is diagnosed if the imputation chains are not freely intermingled with one another, or if the chains show definite trends [@buur18, $\S$ 6.5.2].

@@

A suggestion by @buur18 for a multivariate $\theta$ to monitor is the estimated quantity of scientific interest $\hat{Q}$, which usually is multivariate in nature. This $\theta$ is computed as the estimate in each imputation, $\hat{Q}_\ell$, across iterations. Implementing this, however, might be somewhat too technical for empirical researchers. Besides, this scalar summary is not model-independent. That is, $\theta=\hat{Q}$ is not universal to all complete data problems, while one of the advantages of iterative imputation is that the missing data problem and scientific problem are solved independently. Focusing on the convergence of outcome parameters may influence the iterative imputation procedure in the sense that the model of evaluation favors the model of interest. On these grounds, such a $\theta$ can be considered insufficient too.

@@

<!-- Recommendation: "the different streams should be freely intermingled with one another, without showing any definite trends". -->

<!-- **[MERGE with sections above]** Convergence has historically been inspected visually, by monitoring the imputation algorithm over iterations. This is typically done through traceplots. In a traceplot, the iteration number is plotted against a certain $\theta$. $\theta$s are scalar summaries of the state-space of the algorithm at a specific iteration.  -->

<!-- The default scalar summaries in `mice` are chain means and chain variances. Additionally, researchers may "monitor some statistic of scientific interest" [@buur18, $\S$ 6.5.2]. As @buur18 describes, researchers can specify their quantity of scientific interest $Q$ (e.g., a regression coefficient) as scalar summary $\theta$. Such a user-defined scalar summary, however, may be somewhat advanced for empirical researchers.  -->

@@

<!-- **Since convergence of iterative imputation algorithms is in distribution, there is not a unique point at which the algorithm reaches a converged state. We can only evaluate scalar summaries of the multivariate state-space of the algorithm, $\theta$s. The recommended scalar summaries to evaluate are chain means and chain variances. Additionally, researchers may "monitor some statistic of scientific interest" [@buur18, $\S$ 6.5.2]. Because of the issues mentioned above, we propose a novel $\theta$ to monitor: the first eigenvalue of the variance-covariance matrix of the completed data. Let $\lambda_{1, \ell} \geq \lambda_2 \geq ... \geq \lambda_j$ be the eigenvalues of $\Sigma$ in each of the $m$ completed datasets ${y_{\rm obs}, y_{\rm imp}}$. $\lambda_{1, \ell}$ is measure that summarizes the covariances in the completed datasets.**  -->

@@

<!-- Requirements for diagnostics to identify non-convergence in iterative imputation are: 1) work for a relatively low number of iterations (defaults in popular iterative imputation software between 5 and 30); 2) may be applied to different $\theta$s, including model-independent $\theta$s (e.g., chain mean, or eigenvalue). -->

@@

<!-- A widely used diagnostic to monitor mixing between chains is the potential scale reduction factor $\widehat{R}$ ['Gelman-Rubin statistic'; @gelm92].  Therefore, we use $\widehat{R}$ **[WHICH? Explain that we'll use both]** and $AC$ to evaluate mixing and stationarity separately, as recommended by e.g., @cowl96. -->

<!-- **[Add: we don't use effective sample size and MC error because they assume independent samples *within* chains. In MI, we only use the final estimate. Also, computing n_eff assumes infinitely many iterations, which is fine in Bayesian analyses where T is often at least one thousand. But the default in iterative imp is often substantially lower.]** -->

@@

\begin{align*}
W&=\frac{1}{m} \sum_{\ell=1}^{m} s_{\ell}^{2}, \text { where } s_{\ell}^{2}=\frac{1}{T-1} \sum_{t=1}^{T}\left(\theta^{(t \ell)}-\bar{\theta}^{(\cdot \ell)}\right)^{2}. 
\end{align*}
<!-- %\text{ \cite[p.~5]{veht19}} -->

@@

\begin{align*}
B&=\frac{T}{m-1} \sum_{\ell=1}^{m}\left(\bar{\theta}^{(\cdot \ell)}-\bar{\theta}^{(\cdot \cdot)}\right)^{2}, \text { where } \bar{\theta}^{(\cdot \ell)}=\frac{1}{T} \sum_{t=1}^{T} \theta^{(t \ell)} \text{, } \bar{\theta}^{(\cdot \cdot)}=\frac{1}{m} \sum_{\ell=1}^{m} \bar{\theta}^{(\cdot \ell)}. 
\end{align*}
<!-- %\text{ \cite[p.~5]{veht19}} -->

@@

<!-- close to zero indicate  independence between subsequent $\theta$s in imputation chains.  -->

<!-- Negative $AC$-values show no threat to the stationarity component of convergence. On the contrary even---negative $AC$-values indicate that $\theta$-values of subsequent iterations diverge from one another, which may increase the variance of $\theta$ and speed up convergence. As convergence diagnostic, the interest is therefore in positive $AC$-values.  -->

<!-- Positive $AC$-values occur when $\theta$-values of subsequent iterations are similar. When , we speak of recurrence. Recurrence within imputation chains indicates trending. Ergo, non-stationarity may be diagnosed when $AC$-values are larger than zero. -->

<!-- Negative $AC$-values indicate divergence within imputation chains. **Subsequent sampled values within each imputation chain are less alike.** Positive $AC$-values indicate recurrence. If $\theta$-values of subsequent iterations are similar, trending may occur.  -->

@@

<!-- We assume that complete convergence, defined as $\widehat{R} = 1$ and $AC = 0$, will not be observed.  -->
It is unlikely that iterative algorithms such as MICE will achieve the ideal values of $\widehat{R}=1$ and $AC=0$. Because of its convergence to a distribution, the algorithm will show some signs of non-mixing and non-stationarity even in the most converged state. The aim, therefore, is to reach approximate convergence [@gelm13]. Upon approximate convergence, the imputation chains intermingle such that the only difference between the chains is caused by the randomness induced by the algorithm ($\widehat{R} > 1$), and there is little dependency between subsequent iterations of imputation chains ($AC > 0$). In practice, we diagnose non-convergence when approximate convergence is violated, i.e., when $\widehat{R}$ and $AC$ exceed a certain threshold. The conventional thresholds to diagnose non-mixing are $\widehat{R} > 1.2$ [@gelm92] or $\widehat{R} > 1.1$ [@gelm13]. @veht19 proposed a much more stringent threshold of $\widehat{R} > 1.01$. The magnitude of $AC$-values may be evaluated statistically, using a Wald test with $AC=0$ as null hypothesis [@box15]. $AC$-values that are significantly higher than zero indicate non-stationarity.

<!-- We assess discrepancy from convergence via practical significance (or some conventional version thereof, such as Rh > 1.1) @gelm13, p. 288 -->

<!-- We can use effective sample size ^neff to give us a sense of the precision obtained from our simulations. As we have discussed in Section 10.5, for many purposes it should suffice to have 100 or even 10 independent simulation draws. (If neff = 10, the simulation standard error is increased by p 1 + 1/10 = 1.05). As a default rule, we suggest running the simulation until ^neff is at least 5m, that is, until there are the equivalent of at least 10 independent draws per sequence (recall that m is twice the number of sequences, as we have split each sequence into two parts so that Rb can assess stationarity as well as mixing). Having an effective sample size of 10 per sequence should typically correspond to stability of all the simulated sequences.  @gelm13, p. 287 -->

@@

```{r diagnostics, fig.cap="Convergence diagnostics applied on the two scenarios from Figure 2 ($\\theta$ = chain mean of variable $j$; $m=5$; $T=10$). Depicted are the imputation chains in panel A, two versions of the autocorrelation in panel B, and two versions of $\\widehat{R}$ in panels C and D."}
# load("./images/diagnostics_plot.Rdata") #bquote(theta ~ "= chain mean in "~ y[imp][",m"])
(theta + plot_layout(guides = "collect", ncol = 2)) + (ac_both + plot_layout(guides = "keep")) + old_rhat + new_rhat +  plot_annotation(tag_levels = "A", tag_suffix = ".") # add 'ncol=1' for plots under each other
```

@@

The $\widehat{R}$-values in panel D are conform our expectations. The adapted version of $\widehat{R}$ depicted here does indicate less signs of non-convergence as the number of iterations goes up. An exception to this general trend is the 'dip' around $t=3$. As diagnostic tool, this may lead to the incorrect conclusion that the non-convergence scenario has superior convergence compared to the typical scenario. We can explain this as a downside of the low number of iterations in iterative imputation: the adapted version of $\widehat{R}$ can only be completely employed if the number of iterations is at least four. Otherwise, it is not possible to perform all three transformations to the chains of $\theta$-values. For low values of $t$, the $\widehat{R}$-values conform Vehtari et al. are, therefore, more similar to the original $\widehat{R}$ by Gelman and Rubin.
<!-- However, the superior performance of the typical convergence over the pathological non-convergence is less prominent, even flipped for $t<4$.  -->

@@

<!-- is an unbiased and confidence-valid**[REMOVE?] Subsequently we will evaluate how well $\widehat{R}$ and $AC$ perform in identifying the effects of non-convergence **[and performance of different thetas]**. And finally, we will formulate an informed advice on the requirements to safely conclude sufficient convergence in practice. That is, we conclude that  convergence is sufficient when each estimate $\bar{Q}$ is an unbiased and confidence-valid estimate of the corresponding estimand $Q$.** -->

@@

<!-- **["early stopping" en "proportion of complete cases"]** :  -->

<!-- we vary the missingness proportion $p_{\rm inc}$ in the incomplete data, and we vary the number of iterations $T$ in the imputation algorithm.  -->

@@

<!-- **[REMOVE?] $Q$s are the targets that will be estimated in each simulation repetition, after first *amputing* the complete data with varying missingness proportions, then imputing the missingness with a varying number of iterations, then estimating the $Q$s on the $m$ completed datasets, and finally pooling the $m$ estimates $\hat{Q}_\ell$ to obtain a single $\bar{Q}$ for each $Q$. We try to predict the effect of non-convergence (i.e., the difference between $Q$ and $\bar{Q}$) with the convergence diagnostics $\widehat{R}$ and $AC$, by applying these methods to several scalar summaries of the state-space of the algorithm, $\theta$s.** -->

@@

<!-- **(bruggetje hier)** For multiple imputation algorithms, it holds that convergence is reached when there is no dependency between subsequent iterations of imputation chains ($AC = 0$), and chains intermingle such that the only difference between the chains is caused by the randomness induced by the algorithm ($\widehat{R} = 1$). **We expect that a completely converged state will not be reached.**  -->

<!-- This study evaluates whether $\widehat{R}$ and $AC$ could diagnose convergence of multiple imputation algorithms. We assess the performance of the two convergence diagnostics against the recommended evaluation criteria for MI methods [i.e., average bias, average confidence interval width, and empirical coverage rate across simulations; @buur18, $\S$ 2.5.2]. **That is, there is no baseline measure available to evaluate performance against.** -->

<!-- Based on an empirical finding [@lace07], we hypothesize that $\widehat{R}$ will over-estimate non-convergence of MI algorithms **explain why?**. Lacerda concluded that rhat dropped below 1.2 after at least 50 iterations. Therefore we hypothesize that Vehtari's proposed threshold of $\widehat{R} < 1.01$ will be too stringent for diagnosing convergence in iterative imputation. This over-estimation may, however, be diminished because $\widehat{R}$ can falsely diagnose convergence if initial values of the algorithm are not appropriately over-dispersed [@broo98, p. 437]. In e.g. `mice`, initial values are chosen randomly from the observed data. Therefore, we cannot be certain that the initial values are over-dispersed. We expect this to have little effect on the hypothesized performance of $\widehat{R}$. **(Add actual hypothesis: we thought that mice would converge sooner than $\widehat{R}$ would indicate that it did). On the one hand, $\widehat{R}$ is known to under-estimate signs of non-convergence when the initial values of the algorithm are not appropriately over-dispersed [@broo98, p. 437], which may be the case for iterative imputation algorithms. In e.g. `mice`, initial values are chosen randomly from the observed data. Therefore, we cannot be certain that the initial values are sufficiently over-dispersed. 
**  -->

@@

<!-- Each of these $Q$s is estimated with a corresponding $\bar{Q}$, after the complete dataset is first *amputed* (with a varying missingness proportion $p_{\rm inc}$) and subsequently imputed (with a varying number of iterations $T$).  -->

@@

<!-- In each simulation repetition, we *ampute* the complete dataset five times to obtain five different incomplete datasets, $y$s. The ratio between $y_{\rm obs}$ and $y_{\rm mis}$ in each $y$ depends on the missingness proportion: $p_{\rm inc} =.05,.25,.5,.75,.95$. We ampute the complete data with the `mice` package [function `mice::ampute()`; @mice]. We consider all possible univariate and multivariate patterns of missingness, conform a 'missing completely at random' missingness mechanism [@rubin87]. I.e., the probability to be missing is the same for all $N \times k$ cells in $y$, conditional on the missingness proportion $p_{\rm inc}$.  -->
<!-- <!-- This note only considers a 'missing completely at random' (MCAR) mechanism, where the probability of being missing is equal for all $n \times k$ cells in $y$ [@rubin87]. The probability to be missing is the same for all $n \times k$ cells in $y$ [a 'missing completely at random' missingness mechanism; @rubin87]. -->

<!-- Missing datapoints in each incomplete dataset $y$ are imputed with the `mice()` function [@mice]. All imputation procedures are performed with Bayesian linear regression imputation, and five imputation chains ($m=5$). Because $m=5$, each run of the `mice()` function results in five sets of imputations: $\dot y_{\rm imp, \ell}$, where $m = 1, 2, ..., 5$. The number of iterations in each run of the algorithm varies between simulation conditions ($T=1,2,...,100$).  -->
<!-- <!-- Each repetition of the simulation thus starts from 1 complete dataset, then splits into 5 missingness conditions ($p_{\rm inc} =.05,.25,.5,.75,.95$), each of the amputed datasets is imputed 5 times ($m = 1, 2,..., 5$) for each iteration condition ($t = 1, 2,..., 100$). From these $5\times5\times100$ imputations, we extract the $\theta$s to apply $\widehat{R}$ and $AC$ on (**i.e., chain means, chain variances, $\beta$s per imputation, and the first eigenvalue of the variance-covariance matrix per imputation**).  -->

<!-- From the 5 imputations $\dot y_{\rm imp, \ell}$ we obtain the chain means and chain variances for each variable $j$. These will serve as scalar summary $\theta$ to apply non-convergence diagnostics $\widehat{R}$ and $AC$ on. We then use `mice::complete()` to combine the imputed data $\dot y_{\rm imp, \ell}$ with the observed data $y_{\rm obs}$, resulting in five completed datasets {$y_{\rm obs}, y_{imp, m}$}. -->

<!-- On the 5 completed datasets {$y_{\rm obs}, y_{imp, m}$}, we:  -->

<!-- - Estimate $\hat{Q}_\ell$s for $Q= \mu$ and $Q=\sigma$ with `base::mean()`, and `stats::sd()` [@R]. We pool the $\hat{Q}_\ell$s with `base::mean()` to get $\bar{Q}$s. -->

<!-- - Compute $\lambda_{1, \ell}$, to use as $\theta$ in the two convergence diagnostics. By definition, the first eigenvalue of the variance-covariance matrix, $\lambda_{1, \ell}$, is equal to the variance of the principal component solution of each {$y_{\rm obs}, y_{imp, m}$}, which we calculate with `stats::princomp()` [@R].  -->

<!-- - Perform multiple linear regression with `stats::lm()` [@R] to get the $\hat{Q}_\ell$s for $Q=\beta$ and $Q=r^2$. We pool the $\hat{Q}_\ell$s conform @vink14 to get $\bar{Q}$s for the $\beta$s and $r^2$ **[check  finite pooling of  $r^2$, instead of `mice::pool.r.squared()`]**. Additionally, we use the $\hat{Q}_\ell$s of $Q=\beta$ as $\theta$ to apply the convergence diagnostics to. -->

<!-- <!-- After each run of the `mice` algorithm, the function `mice::complete()` is used to combine the imputed data $\dot y_{\rm imp, \ell}$ with the observed data $y_{\rm obs}$, resulting in five completed datasets {$y_{\rm obs}, y_{imp, m}$}.  -->

<!-- <!-- The estimator for each $Q$ is $\bar{Q}$---the pooled aggregate of the $\hat{Q}_\ell$s across imputations. To calculate the $\bar{Q}$s, we combine the observed data $y_{\rm obs}$ and the imputed data for each imputation $\dot y_{\rm imp, \ell}$ with the function `mice::complete()`. Descriptive statistics are computed as the average across imputations for each variable ($\mu_j$ and $\sigma_j$, where $j = Y, X_1, X_2, X_3$). Estimated regression coefficients are obtained with the function `stats::lm()` for each imputation, and then pooled conform @vink14. The coefficient of determination is estimated for each imputation, and pooled using `mice::pool.r.squared()`.  -->

<!-- We apply non-convergence diagnostics $\widehat{R}$ and $AC$ to each scalar summary $\theta$. We calculate $\widehat{R}$ by implementing Vehtari et al.'s [-@veht19] recommendations, and $AC$ as the correlation between the $t^{th}$ and the $(t+1)^{th}$ iteration.  -->

@@

<!-- Simulation conditions that are impacted by non-convergence portray more extreme bias in the estimated $Q$s, and non-nominal coverages.  -->

@@

<!-- are appropriate,  **[appropriate methods thus identify these conditions]** -->

>>>>

```{r setup, include=FALSE}
# latex options
knitr::opts_chunk$set(fig.width = 4.9, echo = FALSE, fig.align = "center")

# load packages
library(mice) #version 3.8.0
library(dplyr) #version 0.8.5
library(ggplot2) #version 3.3.0
library(patchwork) #version 1.0.0

# load example non-convergence and figures
load("../Results/example_nonconv.Rdata")
load("../Results/example_conv.Rdata")
load("../Results/example_diagnostics.Rdata")
source("../Figures/PlotExample.R")

# load results and figures
load("../Results/complete.Rdata") 
results <- results %>% filter(t<51)
source("../Figures/PlotResults.R")

# default graphing behavior
theme_update(
 plot.title = element_text(hjust = 0.5),
 plot.subtitle = element_text(hjust = 0.5),
 panel.border = element_blank(),
 panel.grid.major = element_blank(),
 panel.grid.minor = element_blank(),
 panel.background = element_blank(),
 axis.line = element_line(colour = "black"),
 legend.key = element_blank(),
 legend.position = "bottom",
 legend.margin = margin(0,0,0,0),
 text = element_text(size = 8),
 title = element_text(size = 8),
 legend.title = element_text(size = 7)
)
```

@@

Figure \ref{fig:diagram} provides an overview of the steps involved with MI. The missing part $y_{\rm mis}$ of an incomplete dataset is imputed $m$ times. This creates $m$ sets of imputed data $\dot y_{\rm imp, \ell}$, where $\ell = 1, 2, ..., m$. The imputed data is then combined with the observed data $y_{\rm obs}$ to create $m$ completed datasets. On each of these datasets the analysis of scientific interest is performed to estimate $Q$: the quantity of scientific interest (e.g., a regression coefficient). Since $Q$ is estimated on each completed dataset, $m$ separate $\hat{Q}_\ell$-values are obtained. Finally, the $\hat{Q}_\ell$-values are combined into a single pooled estimate $\bar{Q}$. The premise of multiple imputation is that $\bar{Q}$ is an unbiased and confidence-valid estimate of the true---but unobserved---scientific estimand $Q$ [@rubin96].
<!-- "A scientific estimand $Q$ is a quantity of scientific interest that we can calculate if we would observe the entire population" [@buur18, par 2.3.1] -->

```{r diagram, out.width="\\linewidth", include=TRUE, fig.cap="Scheme of the main steps in multiple imputation---from an incomplete dataset, to $m=3$ multiply imputed datasets, to $m=3$ estimated quantities of scientific interest $\\hat{Q}$, to a single pooled estimate $\\bar{Q}$."}
knitr::include_graphics("../Figures/DiagramMI.pdf") #adapted from [@buur18] section 1.4.1.
```

A popular method to obtain imputations is to use the 'Multiple Imputation by Chained Equations' algorithm, shorthand 'MICE'[@mice]. With MICE, imputed values are drawn from the posterior predictive distribution of the missing values. The algorithm is named after its iterative nature: a multivariate distribution is obtained by iterating over a sequence of univariate imputations. Iteration, however, also introduces a potential threat to the validity of the imputations: What if the algorithm has not converged? Are the imputations then to be trusted? And can we rely on the inference obtained on the completed data?

@@

<!-- The default number of iterations in common MI software: SPSS = 10 [(link)](https://www.ibm.com/support/knowledgecenter/SSLVMB_24.0.0/spss/mva/syn_multiple_imputation_impute.html), Mplus = 100?? [(link)](https://pdfs.semanticscholar.org/e20e/29e008592cbfbaa567931f74cdfdb5451405.pdf?_ga=2.55354671.54033656.1584698748-527613517.1584698748), Stata = 10 [(link)](https://www.stata.com/manuals13/mi.pdf, p. 139), Amelia = NA, because re-sampling, not convergence [(link)](https://cran.r-project.org/web/packages/Amelia/Amelia.pdf), en MI = 30 [(link)](https://cran.r-project.org/web/packages/mi/mi.pdf). -->

@@

Let $y$ denote an $n \times k$ matrix containing the data values on $k$ variables for all $n$ units in a sample. The data value of unit $i$ ($i = 1, 2, \dots, n$) on variable $j$ ($j = 1, 2, \dots, k$) may be either observed or missing. The number of units $i$ with at least one missing value, divided by the total number of units $n$, is called the proportion of incomplete cases $p_{\rm inc}$ in dataset $y$. The collection of observed data values in $y$ is denoted by $y_{\rm obs}$; the missing part of $y$ is referred to as $y_{\rm mis}$. For each datapoint in $y_{\rm mis}$, we sample $m \times T$ plausible values, where $m$ is the number of imputations ($\ell = 1, 2, \dots, m$) and $T$ is the number of iterations in the imputation algorithm ($t = 1, 2, \dots, T$). The state-space of the algorithm at a certain iteration $t$ may be summarized by a scalar summary $\theta$ (e.g., the average of the imputed values). The collection of $\theta$-values between $t=1$ (summarizing the state-space of the algorithm at initialization) and $t=T$ (summarizing the state-space for the imputed values) will be referred to as an 'imputation chain'.

@@

@@

@@

Currently, the recommended practice for evaluating the convergence of the MICE algorithm is through visual inspection. After running the imputation algorithm for a certain number of iterations, researchers are encouraged to produce traceplots. In a traceplot, a scalar summary of the state-space of the algorithm $\theta$ is plotted against the iteration number, as depicted in Figure \ref{fig:non-conv}. The default scalar summaries to inspect for the MICE algorithm are chain means and chain variances. Non-convergence is diagnosed if the imputation chains are not freely intermingled with one another, or if the chains show definite trends [@buur18, $\S$ 6.5.2]. 
<!-- Recommendation @buur18: "the different streams should be freely intermingled with one another, without showing any definite trends". -->

@@

A suggestion by @buur18 for a multivariate $\theta$ to monitor is the estimated quantity of scientific interest $\hat{Q}$, which usually is multivariate in nature. This $\theta$ is computed as the estimate in each imputation, $\hat{Q}_\ell$, across iterations. Implementing this, however, might be somewhat too technical for empirical researchers. Besides, this scalar summary is not model-independent. That is, $\theta=\hat{Q}$ is not universal to all complete data problems, while one of the advantages of iterative imputation is that the missing data problem and scientific problem are solved independently. Focusing on the convergence of outcome parameters may influence the iterative imputation procedure in the sense that the model of evaluation favors the model of interest. On these grounds, such a $\theta$ can be considered insufficient too. 
<!-- Researchers may "monitor some statistic of scientific interest" [@buur18, $\S$ 6.5.2].  -->

@@

@@

@@

@@

<!-- "We can use effective sample size ^neff to give us a sense of the precision obtained from our simulations. As we have discussed in Section 10.5, for many purposes it should suffice to have 100 or even 10 independent simulation draws. (If neff = 10, the simulation standard error is increased by p 1 + 1/10 = 1.05). As a default rule, we suggest running the simulation until ^neff is at least 5m, that is, until there are the equivalent of at least 10 independent draws per sequence (recall that m is twice the number of sequences, as we have split each sequence into two parts so that Rb can assess stationarity as well as mixing). Having an effective sample size of 10 per sequence should typically correspond to stability of all the simulated sequences."  @gelm13, p. 287 -->

<!-- We don't use effective sample size and MC error because they assume independent samples *within* chains. In MI, we only use the final estimate. Also, computing n_eff assumes infinitely many iterations, which is fine in Bayesian analyses where T is often at least one thousand. But the default in iterative imp is often substantially lower. -->

@@

\begin{align*}
W&=\frac{1}{m} \sum_{\ell=1}^{m} s_{\ell}^{2}, \text { where } s_{\ell}^{2}=\frac{1}{T-1} \sum_{t=1}^{T}\left(\theta^{(t \ell)}-\bar{\theta}^{(\cdot \ell)}\right)^{2}. 
\end{align*}

@@

\begin{align*}
B&=\frac{T}{m-1} \sum_{\ell=1}^{m}\left(\bar{\theta}^{(\cdot \ell)}-\bar{\theta}^{(\cdot \cdot)}\right)^{2}, \text { where } \bar{\theta}^{(\cdot \ell)}=\frac{1}{T} \sum_{t=1}^{T} \theta^{(t \ell)} \text{, } \bar{\theta}^{(\cdot \cdot)}=\frac{1}{m} \sum_{\ell=1}^{m} \bar{\theta}^{(\cdot \ell)}. 
\end{align*}

@@

@@

It is unlikely that iterative algorithms such as MICE will achieve the ideal values of $\widehat{R}=1$ and $AC=0$. Because of its convergence to a distribution, the algorithm will show some signs of non-mixing and non-stationarity even in the most converged state. The aim, therefore, is to reach approximate convergence [@gelm13]. Upon approximate convergence, the imputation chains intermingle such that the only difference between the chains is caused by the randomness induced by the algorithm ($\widehat{R} > 1$), and there is little dependency between subsequent iterations of imputation chains ($AC > 0$). In practice, we diagnose non-convergence when approximate convergence is violated, i.e., when $\widehat{R}$ and $AC$ exceed a certain threshold. The conventional thresholds to diagnose non-mixing are $\widehat{R} > 1.2$ [@gelm92] or $\widehat{R} > 1.1$ [@gelm13]. @veht19 proposed a much more stringent threshold of $\widehat{R} > 1.01$. The magnitude of $AC$-values may be evaluated statistically, using a Wald test with $AC=0$ as null hypothesis [@box15]. $AC$-values that are significantly higher than zero indicate non-stationarity.
<!-- Threshold Rhat > 1.1 @gelm13, p. 288 -->

@@

```{r diagnostics, fig.cap="Convergence diagnostics applied on the two scenarios from Figure 2 ($\\theta$ = chain mean of variable $j$; $m=5$; $T=10$). Depicted are the imputation chains in panel A, two versions of the autocorrelation in panel B, and two versions of $\\widehat{R}$ in panels C and D."}
(theta + plot_layout(guides = "collect", ncol = 2)) + (ac_both + plot_layout(guides = "keep")) + old_rhat + new_rhat +  plot_annotation(tag_levels = "A", tag_suffix = ".") 
```

@@

The $\widehat{R}$-values in panel D are conform our expectations. The adapted version of $\widehat{R}$ depicted here does indicate less signs of non-convergence as the number of iterations goes up. An exception to this general trend is the 'dip' around $t=3$. As diagnostic tool, this may lead to the incorrect conclusion that the non-convergence scenario has superior convergence compared to the typical scenario. We can explain this as a downside of the low number of iterations in iterative imputation: the adapted version of $\widehat{R}$ can only be completely employed if the number of iterations is at least four. Otherwise, it is not possible to perform all three transformations to the chains of $\theta$-values. For low values of $t$, the $\widehat{R}$-values conform Vehtari et al. are, therefore, more similar to the original $\widehat{R}$ by Gelman and Rubin.

@@

@@

@@

@@

@@

@@

@@

@@
